Simulating High-Dimensional Multivariate Data

### 2021-02-24 010-background.Rmd
updated all bigsimr code
moved generated data into separate scripts

```{r ch010-background, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
library(tidyverse)
library(GGally)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

@@

* SP1: Procedure must scale to high dimensions with practical computable times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

<!-- 
The next section describes the motivating data set we use to provide example applications.
your comment -->

@@

```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
d <- 1000
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "data/brca", d, ".rds" )
if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ) )  {
    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
                            data.type = "RNASeq",
                            p = 6L)
    brca1 <- as.data.frame(t(brca0$dat))
    gene_median <- apply(brca1, 2, median)
    gene_median_order <- order(gene_median, decreasing = TRUE)
    brca  <- brca1[,head(gene_median_order, d)]
    writeLines( substring( row.names( brca ), first=1, last= 12 ), './data/examplePatients.txt' ) 
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
```

@@

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(01122021)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

@@

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

GGally::ggpairs( data = small,
                # lower = list( continuous = wrap(lowerfun)),
                upper = list(
                    continuous = wrap('cor', method = "spearman")
                ), 
                ) + theme_bw()
```

@@

Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively. 
For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

where $\Phi(\cdot)$ is the standard normal CDF.
Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

>>>>

```{r ch010-background, include=FALSE, cache=FALSE}
devtools::load_all()
library(tidyverse)
```

@@

* SP1: Procedure must scale to high dimensions with practical compute times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

<!-- 
The next section describes the motivating data set we use to provide example applications.
-->

@@

@@

```{r ch010-realDataTab}
set.seed(01122021)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes, num_genes)

cap <- paste0("mRNA expression for three selected high-expressing genes, ",
              paste(gene_sample, collapse = ", "), 
              ", for the first five patients in the TCGA BRCA data set.")

small_brca <- brca %>%
  select(all_of(gene_sample)) %>%
  head(n = num_patients) %>%
  as_tibble(rownames = "Patient ID") %>%
  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))

small_brca %>%
  knitr::kable(booktabs = TRUE, caption = cap)
```

@@

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
GGally::ggpairs(
  data = small_brca[,-1],
  upper = list(continuous = GGally::wrap('cor', method = "spearman"))
) + theme_bw()
```

@@

Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates. As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively.

For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

where $\Phi(\cdot)$ is the standard normal CDF. Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.

### 2021-02-24 030-bigsimr-package.Rmd
updated all bigsimr code
moved generated data into separate scripts

```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
library(bigsimr)
Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
## Either let JuliaCall find your installation of Julia or set manually
# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
bs <- bigsimr_setup(JULIA_HOME = "/home/alex/julia-1.5.3/bin/", pkg_check = FALSE)
dist <- distributions_setup() # Julia distributions
library(tidyverse)
library(patchwork)
```

@@

The following descrption is directed towards towards `R` users via `bigsimr` R package.
This package provides an interface to the native code written in Julia, registered as the `Bigsimr`Julia package PROVIDE LINK.
In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `PyBigsimr` at URL that interfaces with the Julia `Bigsimr` package.
The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

@@

```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal.", cache=F}

p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") + 
  labs(x = "Temperature")

@@

```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

@@

Next, we configure the input marginals for later input into `rvec`.
The marginal distributions are specified using `Julias`'s `Distributions.jl` package and stored in an `R` vector.

```{r ch030-margins-alist, echo=TRUE, cache=T}
margins <- c(
    dist$Normal(mean(df$Temp), sd(df$Temp)),
    dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
)
```

@@

```{r ch030-cor-convert, echo=TRUE}
## Step 1. Mapping
(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
## Step 2. Check admissibily
is_valid_correlation <- function(rho) {
  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
}
is_valid_correlation(R_X)
```

@@

Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.

your comment -->

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
The R package `rslurm` enables parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
A single run of `bigsimr:rvec` using `rslurm` can be coded as:

@@

<!-- 
Let's evaluate the quality of simulation performance.

your comment -->

@@

```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
res <- readRDS( 'data/rvecRslurmMapRes.rds' )
res
res %>%
    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
    geom_point() +
    geom_abline( slope = 1, intercept = 0 )
```

>>>>

@@

```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
library(tidyverse)
library(patchwork)
library(bigsimr)

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup()
dist <- distributions_setup()
```

The following description is directed towards towards `R` users via the `bigsimr` `R` package. This package provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` at PROVIDE URL that interfaces with the Julia `Bigsimr` package. The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)). When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

@@

```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal."}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") + 
  labs(x = "Temperature")

@@

```{r ch030-aq-temp-pars, echo=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

@@

Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.

```{r ch030-margins-alist, echo=TRUE}
margins <- c(
    dist$Normal(mean(df$Temp), sd(df$Temp)),
    dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
)
```

@@

```{r ch030-cor-convert, echo=TRUE}
# Step 1. Mapping
(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))

# Step 2. Check admissibility
is_valid_correlation <- function(rho) {
  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
}
is_valid_correlation(R_X)
```

@@

Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
-->

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server. The R package `rslurm` enables parallel `rvec` calls. This example assumes that `bigsimr` is installed on a system with a slurm scheduler. A single run of `rvec` using `rslurm` can be coded as:

@@

<!-- 
Let's evaluate the quality of simulation performance.
-->

@@

```{r ch030-rvecRslurmMapFig, eval=FALSE, out.width= '75%', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
res <- readRDS( 'data/rvecRslurmMapRes.rds' )
res
res %>%
    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
    geom_point() +
    geom_abline( slope = 1, intercept = 0 )
```
